The interactions of rational, pragmatic agents  \n lead to efficient language structure and use

### 2019-01-30 paper/paper.Rmd
update with some (not all) of NDGs comments

abstract: 
    Languages display a diverse set of distributional regularities such as the relation between a word's frequency and rank in a corpus, the distribution of dependency lengths, or the presence of lexical properties such as ambiguity. We discuss a framework for studying how these properties emerge from in-the-moment interactions of rational, pragmatic speakers and listeners. Our work takes Zipfian notions of lexicon-level efficiency as a starting point, connecting these ideas to Gricean notions of conversational-level efficiency. To do so, we derive an objective function for measuring the communicative efficiency of linguistic systems and then examining the behavior of this objective in a series of simulations focusing on the communicative function of ambiguity in language. These simulations suggest that rational pragmatic agents will produce communicatively efficient systems and that interactions between such agents provide a framework for examining efficient properties of language more broadly.

@@

But how does such optimization take place? The example given by Zipf [-@Zipf1949a], describes local, communicative interactions in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance, $m \rightarrow u$. Other functionalist projects have also assumed this basic reference game setting [@RegierKempKay2015a; @Piantadosi2011a] and this simplification of the communicative act has proven productive in theoretical [@FerreriCancho2018a], simulation-based [@KirbyGriffithsSmith2014a] and empirical explorations [@HawkinsFrankeSmithGoodman2018a] of efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to speakers' pragmatic reasoning in the moment. Under a Gricean treatment of pragmatics, speakers and listeners are assumed to follow a set of conversational maxims in which they cooperate to transfer information (Grice, 1975). These maxims appear to emerge from efficiency concerns, however (Horn, 1984).\par

In this work, we present a step toward formalizing the framing presented above  -- connecting Zipfian notions of efficient language structure to Gricean notions of rational conversation [@Grice1975a]. To do so minimally requires three basic ingredients -- (1) a language property we’d like to explain, (2) a framework for describing in-the-moment interactions of speaker-listeners, and (3) some measure of linguistic efficiency. For the current project we focus on the “communicative function of ambiguity” for our property -- translating the functionalist theory pursued by Piantadosi, et al. [-@Piantadosi2011a] into our framework. To model local, conversational interactions (2) we adopt the Rational Speech-act framework (RSA) [@FrankGoodman2012a; @GoodmanFrank2016a]. Finally, (3) we derive an objective function for measuring language efficiency in the reference game setting.\par

The current work examines the question, “how does having \textit{lexical ambiguity} make using language $\ell$ more or less useful for communication”? We begin with a high-level introduction to the modelling framework introducing the basic ingredients we will need to represent language as repeated reference games. We derive anobjective function for measuring the efficiency of linguistic systems in this setting. We move on to two case-studies exploring the questions posed above -- examining the efficiency of ambiguity from a language design and language use perspective. The contributions of this work is to show how the reference game framework, in combination with formal modles of communication, can be used to study questions at a range of different timescales, from language use in-the-moment to language change across historical time.\par

@@

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game. We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}. We consider a set of contexts $C$ with an associated prior $P(C)$. Contexts describe different orderings of the underlying meaning prior, formalized as different conditional distributions over meanings e.g. $p(M|C=c)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meaning pairs.\par

@@

An RSA *speaker agent* defines a conditional distribution over utterances, mapping from intended meanings $M$ to utterances $U$ using $\ell$. That is, a speaker defines $P_{speaker}(u|m;\ell)$. We will use $S(u|m;\ell)$ as short-hand throughout.  A *listener agent* defines a conditional distribution over meanings, mapping from utterances $U$ to meanings $M$ using $\ell$. We will use $L(m|u; \ell)$ as shorthand. Note that both speakers and listeners can induce joint distributions over the set of all events $E$, although, importantly, these distributions may differ:
$$P_{speaker}(u, m; \ell) = S(u|m; \ell)p(m)$$
$$P_{listener}(u, m; \ell) = L(m|u; \ell)p(u)$$

@@

Zipf (1949) proposed that the particular distributional properties found in natural language emerge from competing speaker and listener pressures. We operationalize this objective in equation (1) -- the efficiency of a linguistic system $\ell$ being used by speaker and listener agents $S$ and $L$ is the sum of the expected speaker and listener effort to communicate over all possible communicative events $E$.\par

@@

We assume listener effort is the negative log probability of the listener's conditional $-log_2(L(M|U))$. This operationalization of listener effort is intuitively related to existing work in sentence processing in which word difficulty during comprehension is proportional to surprisal [@Hale2001a; @Levy2008a].

@@

Finally, we assume that events $e = <u, m>$ are sampled according the to following generative model -- some object occurs in the world with probability $p(M=m)$, the speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m)$ (i.e. $e \sim p(m)S(u|m)$). Using these basic ingredients we arrive at the following objective:

\begin{equation}
\begin{split}
  = \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(P_{listener}(u, m; \ell))]
\end{split}
\end{equation}
This equation is the simply the cross-entropy between the speaker and listener joint distributions.
\begin{equation}
\begin{split}
  = \mathbb{E}_{P_{speaker}}[-log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener}; \ell)
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive: cross-entropy gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, this is the difference between the joint distribution assumed by the speaker $P_{speaker}$ and listener $P_{listener}$. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listener think. For a complete derivation and its components we refer the reader to Section 2.1 of our supplementary materials.\par

@@

We provide a toy example of their argument here. Say we have two objects in the world $m_1$ and $m_2$, and two languages $\ell_1$ and $\ell_2$. In language $\ell_1$, the low-cost utterance $u_1$ can be used to refer to both $m_1$ and $m_2$ (i.e. $[[u_1]]_{\ell_1} = \{m_1, m_2 \}$). With language $\ell_2$, however, the low cost utterance $u_1$ can only refer to $m_1$ and a higher-cost utterance $u_2$ can refer to $m_2$ (i.e. $[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because she can always use a low-cost utterance), it is more difficult for a listener (because $u_1$ is ambiguous). If context isn't disambiguating then we might prefer $\ell_2$ to $\ell_1$. But, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred.\par

@@

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=8, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only consideration."}
img <- png::readPNG("figs/fig1.png")
grid::grid.raster(img)
```

@@

We adpot the framework above, and (following Piantadosi et al. 2011), add \textit{context} ($C$). The authors argue that mapping two meanings $m_1$ and $m_2$ to a single utterance $u_1$ is useful when they can be disambiguated *in context*. In our case we consider a context $c_i \in C$ to specify a permutation of the meaning probabilities. That is, $p(m|c_1) \neq p(m|c_2)$, when there are two contexts $|C| = \{c_1, c_2\}$.This in turn, leads to an update to our linguistic efficiency objective as we now would like to consider the average speaker-listener effort over all contextualized communicative events (see section 3 SI).\par

@@

Recall that language efficiency is both a function of the particular semantic mappings induced by that language, the speaker and listener agents ($S$ and $L$), as well as the utterance ($P(U)$), meaning ($P(M)$), and context priors ($P(C)$). Rather than assume particular structure for our utterance, meaning, and context prior distributions, for each simulation we generate $P(U) \sim \text{Dir}(1, |U|)$, $P(M) \sim \text{Dir}(1, |M|)$, and $P(C) \sim \text{Dir}(1, |C|)$ where $\text{Dir}(1, k)$ specifies the uniform Dirichlet distribution over a $k$-dimensional probability vector.\par

@@

Following the argument given by Piantadosi et al. [-@Piantadosi2011a], we want to assess the impact of *context* on our objective. To do so we consider four conditions with $n=500$ simulations each (that is, 500 unique sets of $\{P(U), P(M), P(C)\}$. Our first is a *single-context* condition ($|C|=1$) -- there is a only a single context describing $P(M)$. Our second condition contains two-contexts  ($|C| = 2$) -- we consider efficiency under both $P(M|C=c_1)$ as well as $P(M|C=c_2)$. The third and fourth condition correspond accordingly with $|C|=3$ and $|C| = 4$, respectively.\par

@@

\vspace{1em} \fbox{\parbox[b][][c]{7.3cm}{\centering All experiments, data, and analysis code are available in the public repository for the project: (link will be available upon acceptance)}} \vspace{1em}
\noindent

>>>>

abstract: 
    Languages display a diverse set of distributional regularities such as the relation between a word's frequency and rank in a corpus, the distribution of dependency lengths, or the presence of lexical properties such as ambiguity. We discuss a framework for studying how these properties emerge from in-the-moment interactions of rational, pragmatic speakers and listeners. Our work takes notions of lexicon-level efficiency as a starting point, connecting these ideas to notions of conversational-level efficiency. To do so, we derive an objective function for measuring the communicative efficiency of linguistic systems and then examining the behavior of this objective in a series of simulations focusing on the communicative function of ambiguity in language. These simulations suggest that rational pragmatic agents will produce communicatively efficient systems and that interactions between such agents provide a framework for examining efficient properties of language more broadly.

@@

But how does such optimization take place? The example given by Zipf [-@Zipf1949a], describes local, communicative interactions in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance, $m \rightarrow u$. Other functionalist projects have assumed this basic reference game setting [@RegierKempKay2015a; @Piantadosi2011a] and this simplification of the communicative act has proven productive in theoretical [@FerreriCancho2018a], simulation-based [@KirbyGriffithsSmith2014a] and empirical explorations [@HawkinsFrankeSmithGoodman2018a] of efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to speakers' pragmatic reasoning in the moment. Under a Gricean treatment of pragmatics, speakers and listeners are assumed to follow a set of conversational maxims in which they cooperate to transfer information (Grice, 1975). These maxims appear to emerge from efficiency concerns, however (Horn, 1984). A primary target of the current work is to formalize this connection, by providing a framework for deriving system-level efficiencies from local, interaction behavior. Our claim is that to understand an "efficient" property of a system we need to consider *how* that property is *used* efficiently. Additionally, any quantitative description of "efficiency" should be a function of both the system and users.\par

In the current work we provide a case-study for the kind of framing presented above. We choose a property of languages that could, in principle, vary freely, but shows strong regularities across languages. The explanadum is why this regularity holds. We examine lexical ambiguity as our property, extending ideas by Piantadosi et al. [-@Piantadosi2011a]. We define a novel measure of efficiency that depends on the interactional behavior of interlocutors. For our model of speakers and listeners we adopt the Rational Speech Act (RSA) framework, which is nicely attested in experimental data on pragmatic interaction. With these basic ingredients we have a measure of efficiency given language and speaker-listener agents. Finally, we show that the property of interest is prevalent in languages that optimize this measure of efficiency and also how this property is used efficiently during local interaction. The primary contribution of this work is to show how the reference game frameowrk, in combination with formal models of communication, can be used to connect ideas about system-level efficiencies to in-the-moment language use.\par

@@

### Reference games
Zipf’s example of optimal speaker- and listener-languages took the form of a reference game. We adopt that formulation here, assuming these communication games as our basic unit of analysis. In this setting, speakers and listeners are aware of a set of objects $M$, which we will refer to as *meanings* and are knowledgeable about the set of possible signals $U$ (*utterances*) that can be used to refer to a given meaning (see Figure 1). Utterances may have different relative costs, operationalized via a prior over utterances $P(U)$. Similarly, meanings differ in the relative degree to which they need to be talked about, operationalized as a prior over meanings $P(M)$\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}. We consider a set of contexts $C$ with an associated prior $P(C)$. Contexts encode different meaning priors, formalized as separate conditional distributions over meanings e.g. $p(M|C=c_i) \neq p(M|C=c_j)$. Finally, we consider a set of communicative events $e \in E$ where $<u, m> = e$ is a tuple of utterance, meaning pairs. Note that in the case of single context $|C|=1$ we simply have a prior over meanings $p(M)$, a simplification we will use in introducing our objective function.\par

@@

An RSA *speaker agent* defines a conditional distribution over utterances, mapping from intended meanings $M$ to utterances $U$ using $\ell$ in a given context $c$. That is, a speaker defines $P_{speaker}(u|m; \ell)$. We will use $S(u|m; \ell)$ as short-hand throughout.  A *listener agent* defines a conditional distribution over meanings, mapping from utterances $U$ to meanings $M$ using $\ell$ in a given context $c$. We will use $L(m|u; \ell)$ as shorthand. Note that both speakers and listeners can induce joint distributions over the set of all events $E$, although, importantly, these distributions may differ:
$$P_{speaker}(u, m; \ell) = S(u|m; \ell)p(m)$$
$$P_{listener}(u, m; \ell) = L(m|u; \ell)p(u)$$

@@

Zipf (1949) proposed that the particular distributional properties found in natural language emerge from competing speaker and listener pressures. We operationalize this objective in equation (1) -- the efficiency of a linguistic system $\ell$ being used by speaker and listener agents $S$ and $L$ is the sum of the expected speaker and listener effort to communicate over all possible communicative events $E$. Note that to simplify the introduction of this objective we assume there is just a single context ($|C|=1$). In this case $p(C=c)=1$ and there is only distribution over meanings ($p(m|c) = p(m)$). We provide a more detailed description of this derivation in the supplementary materials, both with and without context $C$.\par

@@

We assume listener effort is the negative log probability of the listener's conditional $-log_2(L(m|u))$. This operationalization of listener effort is intuitively related to existing work in sentence processing in which word difficulty during comprehension is proportional to surprisal [@Hale2001a; @Levy2008a].

@@

Importantly, we assume that events $e = <u, m>$ are sampled according the to following generative model -- some object occurs in the world with probability $p(M=m)$, the speaker attempts to refer to that object by sampling from her conditional distribution $S(u|m)$ (i.e. $e \sim p(m)S(u|m)$). From the ingredients highlighted above, it is possible to derive the following objective between the speaker and listener joint distributions (see SI 2.2 for a complete derivation).

\begin{equation}
\begin{split}
  = -\mathbb{E}_{P_{speaker}}[log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener}; \ell)
\end{split}
\end{equation}
From an information-theoretic perspective this objective is intuitive: cross-entropy gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, this is the difference between the joint distribution assumed by the speaker $P_{speaker}$ and listener $P_{listener}$. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

We provide a toy example of their argument here. Say we have two objects in the world $m_1$ and $m_2$, and two languages $\ell_1$ and $\ell_2$. In language $\ell_1$, the low-cost utterance $u_1$ can be used to refer to both $m_1$ and $m_2$ (i.e. $[[u_1]]_{\ell_1} = \{m_1, m_2 \}$ and $[[u_2]]_{\ell_1} = \emptyset$). With language $\ell_2$, however, the low cost utterance $u_1$ can only refer to $m_1$ and a higher-cost utterance $u_2$ can refer to $m_2$ (i.e. $[[u_1]]_{\ell_2} = \{m_1\}$ and $[[u_2]]_{\ell_2} = \{m_2 \}$). While it is cheaper for a speaker to use $\ell_1$ (because she can always use a low-cost utterance), it is more difficult for a listener (because $u_1$ is ambiguous). If context isn't disambiguating then we might prefer $\ell_2$ to $\ell_1$. But, if context is disambiguating then the speaker can use $u_1$ to refer to either $m_1$ or $m_2$ and $\ell_1$ should be preferred.\par

@@

```{r plot-optimal-langs, fig.env = "figure*", fig.pos = "h", fig.width=8, fig.height=2, fig.align = "center", set.cap.width=T, num.cols.cap=2, fig.cap = "Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only objectives."}
img <- png::readPNG("figs/fig1.png")
grid::grid.raster(img)
```

@@

We adopt the framework above, and (following Piantadosi et al. 2011), add \textit{context} ($C$). The authors argue that mapping two meanings $m_1$ and $m_2$ to a single utterance $u_1$ is useful when they can be disambiguated *in context*. In our case we consider a context $c_i \in C$ to specify a permutation of the meaning probabilities. That is, $p(m|c_1) \neq p(m|c_2)$, when there are two contexts $|C| = \{c_1, c_2\}$.This in turn, leads to an update to our linguistic efficiency objective as we now would like to consider the average speaker-listener effort over all contextualized communicative events (see section 3 SI).\par

@@

Recall that language efficiency is both a function of the particular semantic mappings induced by that language, the speaker and listener agents ($S$ and $L$), as well as the utterance ($P(U)$), meaning ($P(M)$), and context priors ($P(C)$). Rather than assume particular structure for our utterance, meaning, and context prior distributions, for each simulation we generate $P(U) \sim \text{Dir}(1, |U|)$, $P(M|C=c) \sim \text{Dir}(1, |M|)$ (a separete conditional distribution over meanings for each context $c$), and $P(C) \sim \text{Dir}(1, |C|)$ where $\text{Dir}(1, k)$ specifies the uniform Dirichlet distribution over a $k$-dimensional probability vector.\par

@@

Following the argument given by Piantadosi et al. [-@Piantadosi2011a], we want to assess the impact of *context* on our objective. To do so we consider four conditions with $n=500$ simulations each (that is, 500 unique sets of $\{P(U), P(M|C), P(C)\}$. Our first is a *single-context* condition ($|C|=1$) -- there is a only a single context describing $P(M)$. Our second condition contains two-contexts  ($|C| = 2$) -- we consider efficiency under both $P(M|C=c_1)$ as well as $P(M|C=c_2)$. The third and fourth condition correspond accordingly with $|C|=3$ and $|C| = 4$, respectively.\par

@@

\vspace{1em} \fbox{\parbox[b][][c]{7.3cm}{\centering All simulation and analysis code are available in the public repository for the project: (link will be available upon acceptance)}} \vspace{1em}
\noindent
